\documentclass[oneside]{amsart}

\usepackage{
  amsmath,
  amssymb,
  amsthm,
  bm, % bold Greek
  extarrows, % xlongrightarrow
  mathrsfs, % mathscr
  mathtools,
  %newtxtext,newtxmath, % Times New Roman
  stmaryrd % llbracket
}
\usepackage[colorlinks,linkcolor=blue]{hyperref}
\usepackage[paperwidth=6.1in,paperheight=9.25in,textwidth=4.6in,textheight=8in]{geometry}
\usepackage[all]{xy}
\DeclareMathOperator{\annihilator}{ann}
\DeclareMathOperator{\automorphism}{Aut}
\DeclareMathOperator{\automorphic}{\mathcal{A}}
\DeclareMathOperator{\character}{X}
\DeclareMathOperator{\extension}{Ext}
\DeclareMathOperator{\formalspectrum}{Spf}
\DeclareMathOperator{\galois}{Gal}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\GSp}{GSp}
\DeclareMathOperator{\h}{H}
\DeclareMathOperator{\isometry}{Iso}
\DeclareMathOperator{\lie}{Lie}
\DeclareMathOperator{\norm}{N}
\DeclareMathOperator{\PSL}{PSL}
\DeclareMathOperator{\representation}{Rep}
\DeclareMathOperator{\restrict}{R}
\DeclareMathOperator{\shimura}{Sh}
\DeclareMathOperator{\SL}{SL}
\DeclareMathOperator{\SO}{SO}
\DeclareMathOperator{\spectrum}{Spec}
\DeclareMathOperator{\symmetric}{sym}
\DeclareMathOperator{\trace}{tr}
\DeclareMathOperator{\volume}{vol}
\newcommand{\cA}{\mathcal{A}}
\newcommand{\cU}{\mathcal{U}}
\newcommand{\cX}{\mathcal{X}}
\newcommand{\dA}{\mathbf{A}}
\newcommand{\dC}{\mathbf{C}}
\newcommand{\dF}{\mathbf{F}}
\newcommand{\dG}{\mathbf{G}}
\newcommand{\dH}{\mathbf{H}}
\newcommand{\dlambda}{{\bm\lambda}}
%\newcommand{\dpi}{{\bm\pi}}
\newcommand{\dR}{\mathbf{R}}
\newcommand{\dQ}{\mathbf{Q}}
\newcommand{\dS}{\mathbf{S}}
\newcommand{\dsigma}{{\bm\sigma}}
\newcommand{\dZ}{\mathbf{Z}}
\newcommand{\eR}{\mathsf{R}}
\newcommand{\fg}{\mathfrak{g}}
\newcommand{\fH}{\mathfrak{H}}
\newcommand{\fk}{\mathfrak{k}}
\newcommand{\fp}{\mathfrak{p}}
\newcommand{\fE}{\mathfrak{E}}
\newcommand{\fT}{\mathfrak{T}}
\newcommand{\fU}{\mathfrak{U}}
\newcommand{\fV}{\mathfrak{V}}
\newcommand{\fW}{\mathfrak{W}}
\newcommand{\fX}{\mathfrak{X}}
\newcommand{\sF}{\mathscr{F}}
\newcommand{\sL}{\mathscr{L}}
\newcommand{\sO}{\mathscr{O}}
\newcommand{\sV}{\mathscr{V}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\abelian}{\mathrm{ab}}
\newcommand{\analytic}{\mathrm{an}}
\newcommand{\arithfrob}{\mathrm{fr}}
\newcommand{\classical}{\mathrm{cl}}
\newcommand{\cusp}{\mathrm{cusp}}
\newcommand{\discrete}{\mathrm{disc}}
\newcommand{\etale}{\textnormal{\'et}}
\newcommand{\finite}{\mathrm{f}}
\newcommand{\hecke}{\mathcal{H}}
\newcommand{\hida}{\mathbf{h}}
\newcommand{\iso}{\xlongrightarrow\sim}
\newcommand{\langlands}[1]{\prescript{\textnormal{\tiny L}}{}{#1}}
\newcommand{\mult}{\mathrm{m}}

\newtheorem*{conjecture*}{Conjecture}
\newtheorem*{definition*}{Definition}

\setcounter{tocdepth}{1}

\title[Representations and deformation theory in arithmetic]{Automorphic representations and deformation theory in arithmetic geometry}
\author{Daniel Miller}
\address{Department of Mathematics, Cornell University, Malott Hall, Ithaca, NY 14853, USA}
\email{dkmiller@math.cornell.edu}
\keywords{Automorphic representation, Shimura variety, deformation theory, modular form, Hecke algebra, Hida theory}
\date{July 2014}
\urladdr{http://www.math.cornell.edu/~dkmiller}
\subjclass[2000]{11F41,11F70,11F80,11G18,13D10,14G35}

\begin{document}
%\fontsize{10.5}{12}\selectfont 


\begin{abstract}
This is a brief expository note, motivated by \cite{m12}, on the analogy 
between the character variety of the fundamental group of a hyperbolic knot, 
and the $p$-ordinary deformation space of a Galois representation. We have 
two goals: 1) Find an arithmetic analogue of the Jones polynomial. 2) 
Generalize the analogy to arbitrary reductive groups, or at least $\GL(n)$. In 
light of these goals, we begin with a lengthy aside in which we introduce 
modular representations and Hida's $p$-ordinary Hecke algebra from the 
perspective of automorphic Galois representations and eigenvarieties. 
\end{abstract}

\maketitle
\tableofcontents





\section{The analogy between knots and primes}

We start by recalling the analogy described in \cite[ch.3-4]{m12}. The 
punchline is: ``a rational prime $p$ should be thought of as a knot in $S^3$.'' 
One sees this via the \'etale topology. Recall that a morphism $f:X\to Y$ of 
schemes is \emph{\'etale} if it is flat and unramified. Equivalently, $f$ is 
flat and for each $y\in Y$, the fiber $X_y$ is the spectrum of a finite 
product of finite separable extensions of $k(y)$. If $X$ is a connected 
noetherian scheme with chosen base point $x\in X$, the category of finite 
\'etale covers of $X$ is canonically equivalent to the category of finite 
sets with continuous action of $\pi_1(X,x)$. Here $\pi_1(X,x)$ is the 
\emph{\'etale fundamental group} of $X$ at $x$, which we will denote $\pi_1(X)$ 
when $x$ is clear. A good reference for all of this is \cite{milneLEC}. To 
save space, we will write $\pi_1(A)$ instead of $\pi_1(\spectrum A)$. 

We should think of the circle $S^1$ as a $K(\dZ,1)$-space. The arithmetic 
analogue is $\spectrum(\dF_q)$ for any prime power $q$. Indeed, as is the case 
for any field, 
$\pi_1(\dF_q)=\galois(\overline{\dF_q}/\dF_q) \simeq \widehat\dZ$, generated by 
the Frobenius $\arithfrob_q(x)=x^q$. The arithmetic analogue of a $3$-manifold 
is $X=\spectrum(O_F)\smallsetminus S$ for any number field $F$ and finite set 
$S$ of primes in $O_F$. Indeed, for constructible sheaves $\sF$ on $X$, there 
is a ``$3$-dimensional duality'' 
$\extension^\bullet(\sF,\dG_\mult) = \h_c^{3-\bullet}(X,\sF)^\vee$ similar to 
the classical Artin-Verdier duality 
$\extension^\bullet(\sL,\dZ) = \h_c^{3-\bullet}(M,\sL)^\vee$ for local systems 
on a $3$-manifold $M$. 

In topology, a knot is an embedding $K:S^1\hookrightarrow S^3$. More generally, 
we could consider any $K(\dZ,1)\hookrightarrow M$, where $M$ is a $3$-manifold. 
The arithmetic analogue is a map $\spectrum(\dF_q) \to \spectrum(O_F)$ coming 
from a prime ideal $\fp\subset O_F$ with residue field $\dF_q$. In topology, 
the standard way of studying a knot $K:S^1\hookrightarrow S^3$ is to consider 
a small tubular neighborhood $V_K$ of $S^1$ in $S^3$. For simplicity, 
we assume $F=\dZ$; then the arithmetic analogue is an ``infinitesimal \'etale 
neighborhood'' of $\spectrum(\dF_p)\hookrightarrow \spectrum(\dZ)$, namely 
$\spectrum(\dZ_p)\to \spectrum(\dZ)$. The \emph{peripheral group} of $K$ is 
$\pi_1(V_K\smallsetminus K)\simeq \dZ^2$, corresponding to 
$\pi_1(\spectrum(\dZ_p)\smallsetminus \spectrum(\dF_p))=\pi_1(\dQ_p)$. Finally, 
the \emph{knot group} is $G_K=\pi_1(S^3\smallsetminus K)$, corresponding to 
$G_{\dQ,p} = \pi_1(\dZ[\frac 1 p])$. The peripheral map 
$\pi_1(V_K\smallsetminus K) \to G_K$ corresponds to the map 
$\pi_1(\dQ_p) \to G_{\dQ,p}$. 

Unfortunately, this analogy is not perfect. While the topological periphery 
group $\pi_1(V_K\smallsetminus K)$ is abelian, free on generators $l,m$ 
($l$ for latitude, $m$ for meridian) and 
$\pi_1(V_K\smallsetminus K) \to \pi_1(V_K)$ has kernel $\langle m\rangle$, the 
arithmetic case is a more complicated. The group $D_p=\pi_1(\dQ_p)$ is a very 
complicated pro-solvable group. It can be written down in terms of generators 
and relations as in \cite[VII \S 5]{nsw08}, but this is not very helpful. The 
group $D_p$ does have a canonical quotient classifying \emph{tame covers} of 
$\spectrum(\dQ_p)$ which has a presentation 
\[
  \pi_1^\mathrm{tame}(\dQ_p) = \langle \sigma,\tau : \sigma \tau \sigma^{-1} = \tau^p\rangle .
\]
We can think of the nonsensical case ``$p=1$'' as being in exact analogy with 
topology. The kernel of $D_p\to\pi_1(\dZ_p)$ is written $I_p$, and (in analogy 
with topology) called the \emph{inertia group} at $p$. 






\section{Automorphic representations}


\subsection{Adeles}

Let $F$ be a number field. If 
$v$ is a place of $F$, write $F_v$ for the completion of $F$ at $v$, and $O_v$ 
for the ring of integers of $F_v$. Write $\dA=\dA_F$ for the ring of adeles of 
$F$; the most important fact about $\dA$ is that it is a locally compact 
$F$-algebra. It can be defined in many ways: 
\begin{itemize}
  \item The topological direct limit $\varinjlim_S\dA(S)$, where $S$ ranges 
    over all finite sets of valuations of $F$, and 
    \[
      \dA(S) = \prod_{v\in S} F_v \times \prod_{v\notin S} O_v .
    \]
  \item The topological tensor product $(\dR\times\widehat\dZ)\otimes F$.
  \item A restricted direct product $\prod_v' (F_v,O_v)$, consisting of those 
    tuples $(a_v)\in \prod_v F_v$ for which $a_v\in O_v$ for almost all $v$.
  \item Via \cite{gs66}, an initial object in the category of locally compact 
    $F$-algebras with no proper open ideals, and with the intersection of all 
    closed maximal ideals being $0$. 
\end{itemize}
Write $\dA_\finite=\widehat\dZ\otimes F$ for the ring of \emph{finite adeles}. 
It is totally disconnected. 

In \cite{c12}, it is shown that there is a unique product-preserving functor 
$(-)(\dA)$ from affine schemes of finite type over $F$ to topological spaces, 
compatible with closed embeddings, for which $(\spectrum F[t])(\dA)=\dA$ with 
its usual topology. In particular, if $G$ is an algebraic group over $F$, the 
abstract group $G(\dA)$ carries the structure of a locally compact topological 
group. As such, it has a unique (up to scalar) Haar measure $\dd g$.  


\subsection{Hecke algebras}

A good reference for this section is \cite{f79}. 

For the remainder, $G$ is a connected reductive group over $F$. Let 
$\fg=\lie G$. There exists 
an open subset of $U\subset \spectrum(O_F)$ and a ``spreading out'' of $G$ to 
a reductive group scheme on $U$. Up to finitely many places, this spreading out 
is well-defined. In particular, for almost all finite places $v$, the group 
$G(O_v)$ is well-defined for almost all $v$. It is a maximal (open) compact 
subgroup of $G(F_v)$. We normalize Haar measures on $G(F_v)$ so that 
$G(O_v)$ has volume $1$, and choose the Haar measure on $G(\dA)$ to be the 
product of measure on each $G(F_v)$. 

Let $v$ be a finite place. Write $\hecke_v$ for the \emph{Hecke 
algebra} consisting of continuous, locally constant, compactly supported 
functions $G(F_v) \to \dQ$. Multiplication is convolution: 
\[
  (f\star g)(x) = \int_{G(F_v)} f(g) g(y^{-1} x)\, \dd y .
\]
Even though this is written as an integral, it is a finite sum over double 
cosets of open compact subgroups of $G(F_v)$, so no analysis is involved. For 
almost all $v$, the algebra $\hecke_v$ comes with a canonical idempotent, 
$e_v=\chi_{G(O_v)}$. Write $\hecke_\finite$ for the restricted tensor product 
(in the sense of \cite[\S 2]{f79}) of the $\hecke_v$ with respect to the 
$e_v$. It is the direct limit $\varinjlim_S \hecke(S)$, where 
$\hecke(S)=\bigotimes_{v\in S} \hecke_v$, and for 
$T\supset S$, the injection $\hecke(S)\to \hecke(T)$ is induced by the 
$e_v$ for $v\in T\smallsetminus S$. We will also think of $\hecke_\finite$ as 
the algebra of locally constant, compactly supported functions $f$ on 
$G(\dA_\finite)$. 

The ring $F_\infty=F\otimes\dR$ is a finite product of copies of $\dR$ and 
$\dC$, so  $G(F_\infty)$ is naturally a Lie group. Fix a maximal compact 
subgroup $K_\infty\subset G(F_\infty)$. The Hecke algebra 
$\hecke_\infty=\hecke_\infty(G)$ is the convolution algebra of $K_\infty$-finite 
distributions on $G(F_\infty)$ with support in $K_\infty$. There is an 
isomorphism 
\[
  U(\fg_\dC)\otimes_{U(\fk_\dC)} M(K_\infty), \iso \hecke_\infty \qquad D\otimes \mu\mapsto D\star \mu ,
\]
where $\fk=\lie(K_\infty)$ and $M(K_\infty)$ is the algebra of measures on 
$K_\infty$. 

The \emph{global Hecke algebra} of $G$ is 
$\hecke=\hecke_\infty\otimes\hecke_\finite$. We will be interested in 
special classes of representations of $\hecke$. 

For a sufficiently large set $S$ of finite places, it makes sense to define 
$e_S\in \hecke_\finite$ to be the characteristic function of 
$\prod_{v\notin S} G(O_v)$. 

An \emph{admissible} representation of $\hecke_\finite$ is an 
$\hecke_\finite$-module $V$ such that for each $v\in V$, there is a finite set 
$S$ of places for which $e_S\cdot v = v$. 


\subsection{Automorphic representations}

A good reference for this section is \cite{bj79}. 

Let $F$, $G$, \ldots be as above. Let $Z$ be the center of $G$, and choose a 
character $\omega:Z(F)\backslash Z(\dA)\to \dC^\times$. Write 
$L^2(G,\omega)$ for the space of measurable functions 
$f:G(F)\backslash G(\dA)\to \dC$ such that 
\begin{align*}
  f(z x) &= \omega(z) f(x) && z\in Z(\dA) \\
  \|f\|^2 &= \int_{G(F)Z(\dA)\backslash G(\dA)} |f(x)|^2\, \dd x < \infty .
\end{align*}
The space $L^2(G,\omega)$ is a representation of $G(\dA)$ in the obvious 
way. Write $L_\discrete^2(G,\omega)$ for the closed subspace generated by all 
irreducible closed subrepresentations. Let 
$\cA(G,\omega)\subset L_\discrete^2(G,\omega)$ be the space of $K$-finite 
vectors. 

[also need $Z(\fg)$-finiteness]

[Don't define $\cA(G,\omega)$ as a subspace of $L^2$ -- the whole thing isn't. 
Instead, just define $\cA(G)$, the space of automorphic forms.]

Then $\cA(G,\omega)$ is naturally a $\hecke$-module, and as such, 
decomposes as a countable direct sum of irreducible representations with finite 
multiplicities: 
\begin{equation}\label{eq:aut-decomp}
  \cA(G,\omega) = \bigoplus_\pi m(\pi) \pi .
\end{equation}
We call the irreducible admissible representations of $\hecke$ appearing in 
\eqref{eq:aut-decomp} \emph{automorphic representations} of $G$. By 
\cite[th.4]{f79}, each automorphic representation $\pi$ decomposes as a 
restricted tensor product $\bigotimes \pi_v$ of irreducible admissible 
representations of the $\hecke_v$. 

In the remainder, we will often pass without comment between admissible 
representations of $G(\dA_\finite)$ and admissible representations of 
$\hecke_\finite$. This is not hard. Suppose $\pi:G(\dA_\finite)\to \GL(V)$ is 
an admissible representation. For $f\in \hecke_\finite$ and $v\in V$, put 
\[
  f \star v = \int_{G(\dA_\finite)} f(x) \pi(x)\cdot v\, \dd x.
\]
This integral is actually a finite sum. Indeed, we can write $f$ as a finite 
sum of scalars multiples of characteristic functions $\chi_{g K}$, where 
$K\subset G(\dA_\finite)$ is open, compact, and fixes $v$. For such a function, 
we see that 
\[
  \chi_{g K}\star v = \int_{G(\dA_\finite)}\chi_{g K}(x) \pi(x) v\, \dd x = \int_K g v\, \dd x = \volume(K) g  v. 
\]
So the action of $\hecke_\finite$ on $V$ makes sense. Going the other way is 
also easy. If $V$ is an admissible $\hecke_\finite$-module and 
$g\in G(\dA_\finite)$, choose open compact normal $K$ such that 
$\chi_K\star v=v$. Inspired by the above, put 
$g v = \volume(K)^{-1} \chi_{g K}\star v$. 


\subsection{Hecke eigensystems and \texorpdfstring{$L$}{L}-functions}

Let $\pi$ be an automorphic representation of $G$ and choose a nonzero vector 
$u$ in $\pi$. For almost all places $v$, the idempotent $e_v=\chi_{G(O_v)}$ in 
$\hecke_v$ fixes $u$ (in this case, we say that $\pi$ is \emph{unramified} at 
$v$). In particular, the action of $\hecke_v$ on $\pi$ factors 
through that of 
\[
  \hecke_v(O_v) = e_v \hecke_v e_v = C_c^\infty(G(O_v)\backslash G(F_v)/G(O_v)) .
\]
Let $S$ be a set of places outside which $e_v$ fixes $u$. Let 
$\hecke(S)=\bigotimes_{v\notin S} \hecke_v(O_v)$. Then $\pi$ is an irreducible 
admissible module over $\hecke(S)\otimes \bigotimes_{v\in S} \hecke_v$. Since 
$\hecke(S)$ is central in this algebra, it must act via a character 
$\chi:\hecke(S)\to \dC$. The system of homomorphisms 
$\{\chi_v:\hecke_v(O_v) \to \dC:v\notin S\}$ is called a \emph{Hecke 
eigensystem}. 

In the case $G=\GL(n)$, Hecke eigensystems have a particularly easy 
description. A character $\chi:\hecke_v(O_v)\to \dC$ is uniquely determined by 
a semisimple conjugacy class $\sigma_v(\chi)\in \GL(n,\dC)$. If 
$\pi=\bigotimes_v \pi_v$ is an automorphic representation of $\GL(n)$, put 
$\sigma_v(\pi) = \sigma(\chi_{\pi_v})$ and (for finite $v$): 
\[
  L_v(s,\pi) = \det\left(1-\norm(v)\cdot \sigma_v(\pi)^{-s}\right)^{-1} .
\]
For $S$ sufficiently large, we can define the \emph{partial $L$-function} of 
$\pi$ as 
\[
  L_S(s,\pi) = \prod_{v\notin S} L_v(s,\pi) .
\]
This has the expected properties including analytic continuation, a functional 
equation\ldots. In the case $G=\GL(n)$, an automorphic representation $\pi$ is 
determined by $L(s,\pi)$. 





\section{Shimura varieties}

For the rest of this note, the reader should keep in mind the example 
$F=\dQ$, $G=\GL(2)$. Many of the definitions work in greater generality, 
but technicalities (which we wish to avoid) multiply endlessly. 


\subsection{Locally symmetric spaces and their cohomology}\label{sec:local-symm}

Classically, one studies representations of a real semisimple group $G$ by 
fixing a maximal compact $K$, setting $X=G/K$, and studying the regular 
representation of $G$ on $C^\infty(\Gamma\backslash X)$ for $\Gamma\subset G$ 
a discrete group. Big examples are the (affine) modular curves $Y_0(n)$, 
coming from $\Gamma_0(n)\subset \SL(2,\dR)$. We will carry out this 
construction adelically. 

Let $G$ be a connected reductive group over $\dQ$. Put 
$X=Z_\infty \backslash G(F_\infty) / K_\infty$. Let 
$K\subset G(\dA_\finite)$ be an open compact subgroup. We define 
\[
  \shimura_K(G) = G(\dQ)\backslash (X\times G(\dA_\finite)) / K .
\]
A priori, this is only a topological space, but the quotient map 
$X\times G(\dA_\finite)/K \to \shimura_K(G)$ gives $\shimura_K(G)$ the 
structure of a Riemannian manifold. Let $(V,\rho)$ be a representation of $G$. 
There is an induced local system $\sV_\rho$ of $F$-vector spaces on 
$\shimura_K(G)$, whose (global) sections are locally constant sections 
$s:X\times G(\dA_\finite)/K \to V$ such that 
$s(\gamma x) = \rho(\gamma) s(x)$ for $\gamma\in G(\dQ)$. 

The cohomology $\h^\bullet(\shimura_K(G),\sV_\rho)$ is naturally an 
admissible $\hecke_\finite$-module. For open compact $C\subset K$, the function 
$\chi_{g C}$ acts via the correspondence 
\[\xymatrix{
  \shimura_K(G) 
    & \ar@{->>}[l] \shimura_{K\cap C}(G) \ar[r]^-{\cdot g} 
    & \shimura_{K\cap g^{-1} C g}(G) \ar@{->>}[r] 
    & \shimura_K(G) .
}\]
There is a standard compactification of $\shimura_K(G)$, namely its 
\emph{Borel-Serre compactification} $\overline\shimura_K(G)$. Define the 
\emph{cuspidal cohomology} to be 
\[
  \h_\cusp^\bullet(\shimura_K(G),\sV_\rho) = \ker\left(\h^\bullet(\overline\shimura_K(G),\sV_\rho) \to \h^\bullet(\partial\shimura_K(G),\sV_\rho)\right) .
\]
The cuspidal cohomology is also an admissible $\hecke_\finite$-module. In 
fact, we have a \emph{generalized Eichler-Shimura isomorphism} \cite[4.1]{s09}. 
\[
  \h_\cusp^\bullet(\shimura_K(G),\sV_\rho) = \bigoplus_{\substack{\pi\in \cA_\cusp(G,\chi_\rho) \\ K\text{-spherical}}}\h^\bullet(\hecke_\infty, \pi_\infty\otimes \rho)\otimes \pi_\finite^K .
\]
The notation $\h^\bullet(\hecke_\infty,-)$ needs explanation. There is a good 
category of admissible $\hecke_\infty$-modules, and $\hom(\dC,-)$ is 
left-exact. Its derived functor is the $(\fg_\infty,K_\infty)$-cohomology 
$\h^\bullet(\hecke_\infty-)$. 


\subsection{Canonical models}\label{sec:can-model}

A good reference for this is \cite{m98}. In \autoref{sec:local-symm} we 
constructed $\shimura_K(G)$ as a Riemannian manifold. It turns out that there 
is a good definition of ``canonical model'' for Shimura varieties over number 
fields, and in that sense, all Shimura varieties $\shimura_K(G)$ have a 
canonical model over a number field called the \emph{reflex field}. (We have 
been intentionally avoiding use of the Shimura datum necessary to define 
$\shimura_K(G)$ in full generality -- the reflex field depends on this.) 

Let $E$ be the reflex field. Not only does the projective system 
$\shimura(G)=\varprojlim \shimura_K(G)$ descend to the reflex field, but the 
action of $G(\dA_\finite)$ via correspondences descends in a canonical way. 
Moreover, in \cite{harris85} it is shown that our local systems $\sV_\rho$ 
descend in a functorial way to $G(\dA_\finite)$-equivariant local systems 
on $\shimura(G)$. 

The main reason we care about this is that if $\shimura_K(G)$ is smooth and 
$E=\dQ$, then general theorems about \'etale cohomology tell us that 
\[
  \h_{\mathrm{sing},c}^\bullet(\shimura_K(G),\sV_\rho) = \h_{\etale,c}^\bullet(\shimura_K(G)_{\overline\dQ},\sV_\rho(\overline{\dQ_l})) 
\]
after choice of an isomorphism $\dC\simeq \overline{\dQ_l}$. The choice of 
an \emph{arithmetic compactification} of $\shimura_K(G)$ lets us extend the 
action of $\hecke_\finite$ to the \'etale cohomology of $\shimura_K(G)$. 





\section{Modular representations}


\subsection{Modular curves}

In this section, algebraic groups, adeles, etc.~will be taken over $\dQ$. 
Let $n\geqslant 1$ be an integer. We define the following congruence subgroup 
of $\SL_2(\dZ)$:
\[
  \Gamma_0(n) = \left\{\begin{pmatrix}a & b \\ c & d \end{pmatrix}\in \SL(2,\dZ) : c\equiv 0\pmod n\right\} .
\]
Let $K_0(n)$ be the induced subgroup of $\GL_2(\dA_\finite)$. Write $Y_0(n)$ 
for the induced locally symmetric space: 
\[
  Y_0(n) = \shimura_{K_0(n)}(\GL_2) = \GL_2(\dQ)\backslash \GL_2(\dA) / Z_\infty K_\infty K_0(n) .
\]
Here $Z_\infty=Z(\GL_d(\dR))$ and $K_\infty=\SO(2)\subset \GL_2(\dR)$. Put 
$X=\GL_2(\dR)^+/Z_\infty K_\infty$. Note that $\GL_2(\dR)^+/Z = \SL_2(\dR)$, so 
\[
  \GL_2(\dR)^+ / Z_\infty K_\infty \iso \fH = \{z\in \dC:\Im z>0\} 
\]
via $\gamma\mapsto \gamma\cdot i$. The strong approximation theorem tells us 
that 
\[
  \GL_2(\dA) = \GL_2(\dQ) \GL_2(\dR) K_0(n),
\]
so the quotient $\shimura_{K_0(n)}(\GL_2)$ is just 
\[
  (\GL_2(\dQ)\cap K_0(n))\backslash \fH = \Gamma_0(n) \backslash \fH = Y_0(n).
\]
This will be a singular complex-analytic orbifold. There are two ways of 
realizing $Y_0(n)$ and its compactication $X_0(n)$ as curves over $\dQ$: 
\begin{enumerate}
  \item Interpret $Y_0(n)$ as a moduli space for elliptic curves with level 
    structure. This moduli problem makes sense over $\dQ$, so $Y_0(n)$ descends 
    in a canonical way to $\dQ$. 
  \item Use the general theory of canonical models of Shimura varieties. 
\end{enumerate}
The former approach generalizes to a special class of Shimura varieties 
consisting of those of \emph{PEL type} (standing for 
\textbf{P}olarization, \textbf{E}ndomorphism, and \textbf{L}evel structure). 
The theory of PEL-type Shimura varieties is interesting and useful, but we 
won't go into it here. 

Instead, note that the space 
\[
  \fH^\pm = Z_\infty \backslash \GL_2(\dR) / \SO_2(\dR) 
\]
can be interpreted as the set of $\GL_2(\dR)$-conjugacy classes of 
homomorphisms $\dS=\restrict_{\dC/\dR}\dG_\mult\to \GL(2)_\dR$ containing 
\[
  h:(x,y)\mapsto \begin{pmatrix} x & y \\ -y & x \end{pmatrix} .
\]
This is all defined over $\dQ$, so the theory of canonical models discussed 
in \autoref{sec:can-model} tells us that if 
$K\subset\GL_2(\dA_\finite)$ is any open compact subgroup, the quotient 
\[
  \shimura_K(\GL_2) = \GL_2(\dQ)\backslash (\fH^\pm\times \GL_2(\dA_\finite))/K
\]
descends to a uniquely determined curve over $\dQ$. Moreover, this curve has a 
well-defined smooth compactification also defined over $\dQ$, so we don't need 
to worry about the difference between minimal and toroidal compactifications. 


\subsection{The Eichler-Shimura construction}

Let $n\geqslant 3$. 
As above, write $Y_0(n)$ for the Shimura variety $\shimura_{K_0(n))}(\GL_2)$, 
and write $X_0(n)$ for its arithmetic compactification. We are interested in the 
cohomology $\h_\cusp^1(X_0(n),\sV_{\symmetric^{k-2}})$.
At the moment, this is just a $\dC$-vector space with an action of 
$\hecke_\finite$. However, in \autoref{sec:local-symm}, there is an automorphic 
decomposition 
\[
  \h_\cusp^1(X_0(n),\sV_{\symmetric^{k-2}}) = \bigoplus_{\substack{\pi\in \automorphic(\GL_2) \\ \Gamma_0(n)\text{-spherical}}} \h^1(\mathfrak{gl}_2,\SO(2),\pi_\infty\otimes \symmetric^{k-2})\otimes \pi_\finite^{\Gamma_0(n)} .
\]
The computation in \cite[\S3.4-3.6]{harder-1987} tells us that 
$\h^1(\mathfrak{gl}_2,\SO(2),\pi_\infty\otimes \symmetric^{k-2})=0$ unless 
$\pi$ is the automorphic representation coming from a weight-$k$ cuspidal 
eigenform of level $n$, in which case 
$\h^1(\mathfrak{gl}_2,\SO(2),\pi_\infty\otimes \symmetric^{k-2})=\dC\oplus \overline\dC$. 
In particular, 
\[
  \h_\cusp^1(X_0(n),\sV_{\symmetric^{k-2}}) = \bigoplus_{f\text{ eigen-cusp}} \dC\oplus \overline{\dC} = S_k(\Gamma_0(n))\oplus \overline{S_k(\Gamma_0(n))}.
\]

Recall that our modular curves are defined over $\dQ$. So we can consider the 
cohomology spaces 
\[
  \h_{\etale,\cusp}^1(X_0(n)_{\overline\dQ},\overline{\dQ_l}) \simeq S_2(\Gamma_0(n),\dC) .
\]
These have commuting actions of 
$G_{\dQ,l n}=\pi_1(\dZ[\frac{1}{l n}])$ and 
$\hecke_\finite$. So $G_{\dQ,l n}$ acts on each $\hecke_\finite$-irreducible 
piece. These pieces are 2-dimensional, so we get, for each cuspidal eigenform 
$f$, a Galois representation 
$\rho_{f,l}:G_{\dQ,nl} \to \GL_2(\overline{\dQ_l})$. 
The \emph{Eichler-Shimura relation} basically tells us that the Hecke and 
Frobenius parameters for $\pi_f$ and $\rho_{f,l}$ match up. That is, for all 
$p\nmid l n$, we have 
$\rho_{f,l}(\arithfrob_p) = \sigma_p(\pi_f)$, or equivalently 
$L_p(\rho_{f,l},s) = L_p(\pi_f,s)$. 

It is known that $\rho_{f,l}:G_{\dQ,l n} \to \GL_2(\overline{\dQ_l})$ 
factors through $\GL_2(K_{f,\lambda})$, where 
$K_f = \dQ(a_p(f):p\text{ prime})$ is a number field and $\lambda$ is a place 
of $K_f$ dividing $l$. An elementary argument shows that we can conjugate 
the image of $\rho_{f,l}$ to lie in $\GL_2(O_{f,\lambda})$, where 
$O_f=O_{K_f}$. In particular, we can reduce $\rho_{f,l}$ modulo $\lambda$ 
to get a continuous representation 
\[
  \bar\rho_{f,l}:G_{\dQ,l n} \to \GL_2(O_{f,\lambda}/\lambda) = \GL_2(\dF_\lambda) .
\]
We say that a mod-$l$ representation of $G_\dQ$ is \emph{modular} (better, 
\emph{automorphic}) if it is of the form $\bar\rho_{f,l}$ for some $f$. 
Similarly, we say that an $l$-adic representation of $G_\dQ$ is 
\emph{modular} (better, \emph{automorphic}) if it is of the form 
$\rho_{f,l}$ for some $l$. 

In what follows, we will ignore the modular form $f$ and just write $\pi$ for 
the corresponding cuspidal automorphic representation of $\GL(2)$, keeping in 
mind that for some automorphic representations (those corresponding to Maass 
forms) we still don't know how to construct the associated Galois 
representations. For an automorphic representation $\pi$, write 
$\rho_{\pi,l}$ for the corresponding $l$-adic representation (assuming 
it exists). 


\subsection{Interpolating modular representations}

The Hida families we will see later on are essentially $p$-adic families of 
cuspidal automorphic representations of $\GL(2)$. Even better, Hida 
constructed the corresponding $p$-adic family of Galois representations. Here 
we will give the (conjectural) bigger picture. 

Let $G$ be a connected reductive group over $\dQ$. Recall we have a Hecke 
algebra $\hecke=\hecke_\finite\otimes \hecke_\infty$, and automorphic 
representations of $G$ are, by definition, a special class of irreducible 
representations of $\hecke$. Remember that 
$\hecke_\finite$ is the convolution algebra of locally constant, compactly 
supported functions on $G(\dA_\finite)$. If $K\subset G(\dA_\finite)$ is 
open compact, then $e_K=\frac{1}{\mu(K)}\chi_K$ is an idempotent in 
$\hecke_\finite$, and we write $\hecke(K)=e_K \hecke_\finite e_K$ for algebra 
of locally constant, compactly supported, $K$-bi-invariant functions on 
$G(\dA_\finite)$. If $e_K$ acts trivially on an automorphic representation 
$\pi$, we say $\pi$ is \emph{$K$-spherical}. 

The basic idea of a $p$-adic family of automorphic representations is that 
we should fix $K^p\subset G(\dA^p_\finite)$ (the \emph{tame level}) and consider 
families of automorphic representations that are $K^p K_p$-spherical, where $K_p$ 
ranges over a special class of subgroups of $G(\dQ_p)$. One makes this rigorous 
via a modified Hecke algebra. Here we mainly follow \cite[4.1]{urban-2011}. 
Fix a prime $p$ and assume $G$ is split at $p$. Let $(T,B)$ be a Borel pair, 
and let $N$ be the unipotent radical of $B$. Define 
\begin{align*}
  I_r &= \{g\in G(\dZ_p):\bar g\in B(\dZ/p^r)\} \\
  T^- &= \{t\in T(\dQ_p):t N(\dZ_p) t^{-1} \subset N(\dZ_p)\} \\
  \Delta_r^- &= I_r T^- I_r .
\end{align*}
There is an isomorphism 
$u:\dZ_p[T^-/T(\dZ_p)]\to C_c^\infty(I_r \backslash \Delta_r^- / I_r)$ by 
$t\mapsto u_t=\chi_{I_r t I_r}$. Here it is crucial that we normalize the 
Haar measure so that $I_r$ has volume $1$. So we call 
$\cU_p=\dZ_p[T^-/T(\dZ_p)]$ and think of $\cU_p$ as a single avatar for all 
of the $C_c^\infty(I_r \backslash \Delta_r^- / I_r,\dZ_p)$. Our big Hecke 
algebra is 
\[
  \hida = C_c^\infty(K^p\backslash G(\dA_\finite^p)/K_p,\dQ_p)\otimes \cU_p .
\]
Note that $\hida\hookrightarrow \hecke_\finite\otimes \dQ_p$. So, if 
$\sigma$ is an irreducible representation of $\hida$, we will call $\sigma$ 
\emph{automorphic} if $\sigma=\pi_\finite^{K^p I_r}|_\hida$ for some honest 
automorphic representation $\pi$. Note that if $G\ne \GL(n)$, the 
representation $\pi$ may be far from unique. 

A $p$-adic family of automorphic representations of $G$ will contain 
$K^p I_r$-spherical representations for varying $r$. We will also require the 
weight to vary $p$-adically. So, let $\fW$ be the $p$-adic weight space 
determined by 
\[
  \fW(A) = \hom_\mathsf{cts}(T(\dZ_p),A^\times) .
\]
Let $\fW^\classical=\character^\ast(T)$ be the set of \emph{classical 
weights}. We say an automorphic representation $\pi$ is \emph{cohomological of 
weight $\lambda$} if $\pi$ appears in some 
$\h_\cusp^\bullet(\shimura_{K^p I_r}(G),\sV_\lambda(\dC))$. Given such a 
representation, we have the trace map $\trace_\pi:\hida\to \dQ_p$, which 
characterizes $\pi$ (as an $\hida$-module). 

\begin{definition*}
A \emph{$p$-adic automorphic representation $\dsigma$ of $G$ of tame level 
$K^p$} consists of :
\begin{itemize}
  \item An rigid subset $\fU\subset \fW$. 
  \item A finite flat surjection $w:\fT\to \fU$. 
  \item A $\dQ_p$-linear map $J:\hida\to \sO(\fT)$.
\end{itemize}
We require that there is a dense 
subset of $\sigma\mapsto\lambda\in\fU^\classical$ for which $\lambda$ is 
dominant and the composite 
\[
  J_\sigma:\hida \to \sO(\fT) \xrightarrow{\mathrm{ev}_\sigma} \dC_p
\]
is $m(\pi,\lambda) \trace_\pi$ for a cohomological representation $\pi$ of 
weight $\lambda$. 
\end{definition*}
Here $m(\pi,\lambda)$ is the \emph{Euler-Poincar\'e characteristic} 
by 
\[
  m(\pi,\lambda) = \sum (-1)^i \dim\hom_\hida(\pi^{K^p I_r},\h_\cusp^i(\shimura_{K^p I_r}(G),\sV_\lambda(\dC)) .
\]
In \cite{urban-2011}, Urban constructed $p$-adic families containing the 
``finite slope'' representations, for groups satisfying the Harish-Chandra 
condition. 

Recall that for ``nice'' automorphic representations $\pi$, there should 
be a Galois representation 
$\rho_{\pi,p}:G_\dQ\to \langlands G(\overline{\dQ_p})$ unramified almost 
everywhere, such that for unramified $l$, 
$\rho_{\pi,p}(\arithfrob_l)$ is conjugate to the Satake parameter of $\pi$ 
at $l$. For $G=\GL(n)$, this will just be a representation 
$G_\dQ\to \GL_n(\overline{\dQ_p})$. 

\begin{conjecture*}
Let $\dsigma=(\fU,\fV,w,J)$ be an $L$-algebraic $p$-adic automorphic 
representation. Then there is a continuous representation 
$\rho_\dsigma:G_\dQ\to \langlands G(\sO(\fV))$ such whenever 
$\sigma\mapsto \lambda\in \fU^\classical$ is automorphic, the composite 
\[
  \rho_\sigma:G_\dQ\to \langlands G(\sO(\fV)) \xrightarrow{\mathrm{ev}_\sigma} \langlands G(\overline{\dQ_p}) 
\]
is the corresponding $p$-adic Galois representation. 
\end{conjecture*}

This conjecture is \emph{wide} open -- we don't even know how to construct 
individual $\rho_{\pi,p}$ for most $G$. However, for $G=\GL(2)$ and 
$\pi$ corresponding to a $p$-ordinary form, 





\section{Deformation theory}


\subsection{Motivation}

First let's consider the 
motivation for studying deformations of Galois representations. If $X$ is a 
nice (that is smooth, projective and geometrically integral) variety over 
$\dQ$, its \'etale cohomology 
$\h_\etale^\bullet(X_{\overline\dQ},\dQ_l)$ carries a continuous 
action of $G_\dQ$. The ``right'' way to see this is as follows. Spread out 
$X$ to a smooth proper scheme $\cX$ over an open 
$U= \spectrum(\dZ)\smallsetminus S$. Write $\pi:\cX\to U$ for the structure 
map. Then $\eR^\bullet \pi_\ast \dQ_l$ is a local system on 
$U_\etale$. The \'etale version of covering space theory tells us that local 
systems correspond to representations of $\pi_1(U)=G_{\dQ,S}$. 

The motivating example was the representation $\rho_{E,l}$ coming from an 
elliptic curve $E\xrightarrow\pi U$ via $\eR^1 \pi_\ast \dQ_l$. Langlands' 
conjectural framework tells us that there should exist an automorphic cuspidal 
representation $\pi$ of $\GL(2)$ for which 
$\rho_{E,l} \simeq \rho_{\pi,l}$. We don't know how to construct $\pi$ 
this directly. However, we \emph{do} know (via Serre's conjecture) that 
$\bar\rho_{E,l}$ is automorphic. One of the main applications of 
deformation theory is to prove that the automorphy of $\bar\rho_{E,l}$ 
implies that of $\rho_{E,l}$. 

More generally, given an $l$-adic Galois 
representation $\rho:G_{\dQ,S}\to \GL_n(\dQ_l)$ that is suitably nice 
(\emph{geometric}, in the sense of Fontaine-Mazur \cite{fontain-mazur-1995}), 
Langlands' program tells us that we should expect there to be a cuspidal 
automorphic representation $\pi$ of $\GL(n)$ such that 
$\rho\simeq \rho_{\pi,l}$ (assuming we knew how to construct $\rho_\pi$ in 
general). Proving that $\rho$ is automorphic is very hard! However, we have a 
much better chance (in theory and in practice) of showing that 
$\bar\rho:G_{\dQ,S} \to \GL_n(\dF_l)$ is automorphic. A theorem to the 
effect that ``$\bar\rho$ automorphic $\Rightarrow\rho$ automorphic'' is known 
as a \emph{automorphy lifting theorem}. In practice, one has to impose many 
technical conditions on $\bar\rho$ and the automorphic representation with 
$\bar\rho_\pi\simeq \bar\rho$, and one uses groups like $\GSp(n)$ instead of 
$\GL(n)$. 


\subsection{Representations of knot groups}

Our exposition here follows that of \cite[ch.13-14]{m12}. Let 
$K\subset S^3$ be a hyperbolic knot, $M=S^3\smallsetminus K$ the knot 
complement, $\pi=\pi_1(M)$ the knot group. The uniformization $\dH^3\to M$ 
induces a representation $\pi \to \automorphism(\dH^3) = \PSL_2(\dC)$ which 
lifts to $\rho:\pi \to \SL_2(\dC)$. Introduce the \emph{representation 
variety} $\representation(\pi,\SL_2)$ of homomorphisms $\pi\to \SL_2(\dC)$. 
There are two ways of describing $\representation(\pi,\SL_2)$. One elementary 
but useful approach is to write 
$\pi=\langle g_1,\dots,g_m|r_1,\dots,r_n\rangle$. The variety 
$\representation(\pi,\SL_2)$ is just the subset of $\SL_2(\dC)^m$ cut out by 
the relations $r_1,\dots,r_n$. A more functorial definition is to require that 
for all $\dC$-algebras $A$, a natural isomorphism 
\[
  \hom_\mathsf{grp}(\pi,\SL_2(A)) \simeq \hom_{\mathsf{sch}/\dC}(\spectrum A,\representation(\pi,\SL_2)) .
\]

The \emph{character variety} of $K$ is the geometric quotient 
\[
  X_K = \representation(\pi,\SL_2) /\!\!/ \SL_2 = \spectrum\left(\dC[\representation(\pi,\SL_2)]^{\SL_2(\dC)}\right) ,
\]
via the obvious action of $\SL(2)$ on $\representation(\pi,\SL_2)$ via 
conjugation. The representation $\rho$ is a point in $X_K$, and one is 
interested in the connected component $X_K(\rho)$. 


\subsection{Deformation functors}

The analogous situation in number theory is much more complicated, partly 
because $G_S=\pi_1(\spectrum(\dZ)\smallsetminus S)$ is not a finitely presented 
group -- it's a compact topological group which is (conjecturally) 
\emph{topologically} finitely presented. So instead of looking for 
representations $G_S\to \GL_2(\dC)$, we should look for continuous 
representations $G_S\to \GL_2(A)$, where $A$ is a topological $\dZ_p$-algebra. 

Briefly, a scheme $X$ over $k$ can be thought of in terms of its functor of 
points $X(-):k\textnormal{-}\mathsf{Alg}\to \mathsf{Set}$. In the arithmetic 
context, our deformation spaces will be \emph{formal schemes} over $\dZ_p$. For 
us, this just means that the test category consists of complete local 
pro-artinian $\dZ_p$-algebras with residue field $\dF_p$. If $R$ is such an 
ring, we write $\formalspectrum(R)$ to denote the functor 
$A\mapsto\hom(R,A)$. There is a way of making $\formalspectrum(R)$ into a 
topological space with structure sheaf, but we will not need this. 

The functorial approach to defining representation schemes works well. 
If $\pi$ is an arbitrary profinite group, there is a 
formal scheme $\widehat\representation(\pi,\GL_n)$, satisfying 
\[
  \widehat\representation(\pi,\GL_n)(A) = \hom_\mathrm{cts}(\pi,\GL_n A) ,
\]
for any local pro-artinian $\dZ_p$-algebra $A$ with residue field $\dF_p$. The 
problem is, $\widehat\representation(\pi,\GL_n)$ is really horrible as a space 
-- it generally has infinitely many different connected components. So before 
we do anything else, let's restrict to the connected component of $\bar\rho$, 
where $\bar\rho:\pi\to \GL_n(\dF_p)$ has been fixed beforehand. The component 
$\fX^\square(\bar\rho)$ represents continuous homomorphisms 
$\pi\to \GL_n(A)$ that reduce to $\bar\rho$ modulo $p$. 

As before, we will quotient out by the natural action of $\GL(n)$, but here we 
should be careful because $\GL(n)$ does not preserve the component 
$\fX^\square(\bar\rho)$. The correct thing to do is to first define 
\[
  \widehat\GL_n(A) = \{g\in \GL_n(A):g\equiv 1\pmod p)\} = \ker\left(\GL_n(A)\to \GL_n(\dF_p)\right).
\]
The action of $\widehat\GL(n)$ on $\widehat\representation(\pi,\GL_n)$ 
preserves $\fX^\square(\bar\rho)$. Now a miracle happens. Define 
\[
  \fX(\bar\rho)(A) = \fX^\square(A)/\widehat\GL_n(A).
\]
Then in \cite[pr.1]{mazur-1989}, it is proved that if $\bar\rho$ is absolutely 
irreducible and $\pi$ satisfies a certain technical hypothesis (which will hold 
for all our examples), the functor $\fX(\bar\rho)$ is represented by 
a complete local noetherian $\dZ_p$-algebra $R_{\bar\rho}$ with residue field 
$\dF_p$. That is, there is a representation $\rho:\pi\to \GL_n(R_{\bar\rho})$ 
lifting $\bar\rho$ such that any $\widehat\GL_n(A)$-equivalence class of lifts 
$\pi\to \GL_n(A)$ is induced by a unique continuous homomorphism 
$R_{\bar\rho} \to A$. 

Suppose we have fixed a subgroup $I\subset \pi$. We call a representation 
$\rho:\pi\to \GL_2(A)$ \emph{$I$-ordinary} if $\rho^I$ is a free, rank-one, 
direct summand of $\rho$. Suppose $\bar\rho$ is absolutely irreducible and 
$I$-ordinary. Then we can define a subfunctor $\fX^\circ(\bar\rho)$
of $\fX(\bar\rho)$ by 
\[
  \fX^\circ(\bar\rho)(A) = \{\rho\in \fX(\bar\rho)(A):\rho\text{ is $I$-ordinary}\} .
\]
By \cite[pr.3]{mazur-1989}, $\fX^\circ(\bar\rho)$ is represented by a complete 
local noetherian $\dZ_p$-algebra $R_{\bar\rho}^\circ$ with residue field 
$\dF_p$. 


\subsection{The case \texorpdfstring{$n=1$}{n=1}}

The easiest example is when our representations take values in $\GL(1)$. Let 
$\pi=\pi_1(\dZ[\frac 1 p])$ and 
$\bar\rho=\bar\kappa:\pi\to \GL_1(\dF_p)$ be the \emph{mod-$p$ cyclotomic 
character}. This is defined, for $\sigma\in G_\dQ$, by 
\[
  \sigma(\zeta_p) = \zeta_p^{\bar\kappa(\sigma)} .
\]
Let's start by computing $\widehat\representation(\pi,\GL_1)$. Deformations 
$\rho:\pi\to A^\times$ factor through $\pi^\abelian$. Class field theory tells 
us that $\pi^\abelian \simeq \dZ_p^\times$. So 
\[
  \widehat\representation(\pi,\GL_1) = \formalspectrum\left(\dZ_p\left\llbracket \dZ_p^\times\right\rrbracket\right) \simeq \coprod_{\varepsilon:\pi \to \dF_p^\times} \formalspectrum\left(\dZ_p\llbracket \dZ_p\rrbracket\right) ,
\]
where each $\formalspectrum(\dZ_p\llbracket \dZ_p\rrbracket)$ is the connected 
component of some $\varepsilon$. We see that 
$R_{\bar\kappa}\simeq \dZ_p\llbracket \dZ_p\rrbracket\simeq \dZ_p\llbracket X\rrbracket$, 
via $[p]\leftrightarrow X+1$. 


\subsection{The main example}

Let $U\subset\spectrum(\dZ)$ be open, and put $\pi=\pi_1(U)$. Let 
$E\xrightarrow e U$ be an elliptic curve. Choose a prime $p\notin U$. The 
$p$-torsion subscheme $E[p]$ is an \'etale cover of $U$, so we get an action 
of $\pi$ on the underlying set $E[p]\simeq (\dZ/p)^2$. This action preserves 
the group structure, so we have a representation 
\[
  \bar\rho=\bar\rho_{E,p}:\pi_1(U)\to \GL_2(\dF_p) .
\]
Another way of constructing $\bar\rho$ is to use the equivalence between 
\'etale-local $\dF_p$-systems on $U$ and $\dF_p$-representations of $\pi$ to 
realize $\eR e_\ast \dF_p$ as a mod-$p$ representation of $\pi$. 

We could consider the universal deformation ring $R_{\bar\rho}$ and its 
associated formal spectrum $\fX(\bar\rho)$. Recall that if 
$K\hookrightarrow S^3$ is a hyperbolic knot, the ``knot decomposition group'' 
$D_K=\pi_1(\mathrm{torus})$ is abelian, so 
\[
  \rho_K|_{D_K} \sim \begin{pmatrix} \varepsilon & \ast \\ & \varepsilon^{-1} \end{pmatrix} ,
\]
for some character $\varepsilon:G_K\to \dC^\times$. In particular, 
$\varepsilon(I_K)=1$. So to make the analogy between knots and primes more 
precise, we should require that our residual Galois representation 
$\bar\rho:\pi \to \GL_2(\dF_p)$ be \emph{$p$-ordinary} in the sense that 
\[
  \bar\rho|_{D_p} \sim \begin{pmatrix} \varphi & \ast \\ & \psi \end{pmatrix} ,
\]
where $\psi({I_p})=1$. This is exactly ``$I_p$-ordinary'' as defined above. So 
there is a $p$-ordinary representation 
$\rho^\circ:\pi \to \GL_2(R_{\bar\rho}^\circ)$ that is universal for 
$p$-ordinary deformations, 
i.e.~$\formalspectrum(R_{\bar\rho}^\circ)$ represents the 
functor
\[
  \fX^\circ(\bar\rho)(A) = \{\rho\in \fX(\bar\rho)(A):\rho\text{ is $p$-ordinary}\} .
\]





\section{Deformations of hyperbolic structures and Hida theory}

We put together all the machinery we've developed to see an analogy between the 
space of hyperbolic structures on a 3-manifold and the formal spectrum of 
Hida's big Hecke algebras. 


\subsection{Deformation of hyperbolic structures}

Let $K$ be a hyperbolic knot, $M=S^3\smallsetminus K$ the knot complement. Put 
$\Gamma=\pi_1(M)$. Let $T_K=\operatorname{Teich}(\Gamma)$ be the space of 
injective homomorphisms $\Gamma\to \isometry(\dH^3)$ with discrete image, taken 
up to conjugacy; this is the space of hyperbolic structures on $M$. Since 
$\isometry(\dH^3)=\PSL_2(\dC)$, we get a map 
\[
  \phi:T_K \to X_K .
\]
It's not quite straightforward -- you have to fudge for this to be well defined. 
On a small neighborhood, $T_K^\circ \to X_K^\circ$ is an isomorphism. 

Also, $X_K^\circ\to \dC$, $\rho\mapsto \trace \rho(m)$ is locally 
biholomorphic. Similarly, $\fT^\circ(\bar\rho) \to \fX^\circ(\bar\rho)$, 
choose a generator $\tau$ of the $\dZ_p$-quotient of $\pi_1(\dZ[\frac 1 p])$. 
Then $\rho\mapsto \trace\rho$ is $p$-adically bianalytic near $\bar\rho$. 


\subsection{Hida theory}

We showed explicitly how to construct the Galois representations associated to 
cuspidal eigenforms of weight $2$. In fact, in \cite{deligne-1973}, Deligne 
showed how to construct $\rho_{f,l}$ for \emph{any} cusp-eigenform $f$ of 
weight $k\geqslant 2$. Recall that such forms have a Fourier expansion 
\[
  f(z) = \sum_{n\geqslant 1} a_n(f) e^{2\pi n z} .
\]
Say that $f$ is \emph{$p$-ordinary} if $a_p(f)$ is a $p$-adic unit. This is equivalent 
to $\bar\rho_{f,p}$ being an extension of an unramified character by a 
character. In \cite{hida-1986a,hida-1986b}, Hida $p$-adically interpolated the 
Galois representations $\rho_{f,p}$ coming from $p$-ordinary $f$ of varying 
weight and level. 

Fix a prime $p\geqslant 5$ and an integer $n$ prime to $p$. Let $f$ be a 
$p$-ordinary modular form of level $n p^r$. Then there is a $p$-adic 
family $\bm f$ of automorphic representations containing $f$. In fact, Hida 
\emph{explicitly} constructs a completion $\hida^\circ$ of $\hida$ such that 
for $\fT^\circ=\formalspectrum(\hida^\circ)$, \emph{all} $p$-ordinary forms 
of tame level $n$ which are congruent to $f$ lie in $\fT^\circ$. Moreover, 
there is the associated Galois representation 
$\rho_{\bm f}:G_\dQ\to \GL_2(\hida^\circ)$, such that \emph{all} $p$-ordinary 
modular Galois representations congruent to $\rho_f$ come from 
$\hida^\circ$. 





\subsection{The analogy}

Let $f$ be a $p$-ordinary cuspidal eigenform. Put $\rho=\rho_{f,p}$. Let 
$\fT^\circ=\formalspectrum(\hida^\circ)$ be the corresponding $p$-adic family 
of modular forms, and $\rho_{\bm f}:G_\dQ\to \GL_2(\hida^\circ)$ the Galois 
representation. Since $\rho_{\bm f}\equiv \bar\rho_{f,p} \pmod p$ and 
$\rho_{\bm f}$ is $p$-ordinary, we get a map 
$R_{\bar\rho_f}^\circ \to \hida^\circ$, equivalently 
\[
  \phi:\fT^\circ \to \fX^\circ(\bar\rho) .
\]
The \emph{very} hard theorem of Wiles, etc.~is that $\phi$ is an isomorphism if 
$f$ satisfies some technical hypotheses. 





\bibliographystyle{alpha}
\bibliography{tidbit-sources}

\end{document}
