\documentclass{amsart}

\usepackage{amsmath,amssymb,amsthm,bm,fullpage,stmaryrd}
\usepackage[all]{xy}
\usepackage[colorlinks,linkcolor=blue]{hyperref}
\DeclareMathOperator{\adjoint}{ad}
\DeclareMathOperator{\galois}{Gal}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\h}{H}
\DeclareMathOperator{\trace}{tr}
\DeclareMathOperator{\witt}{W}
\newcommand{\cO}{\mathcal{O}}
\newcommand{\dG}{\mathbf{G}}
\newcommand{\dF}{\mathbf{F}}
\newcommand{\dQ}{\mathbf{Q}}
\newcommand{\dT}{\mathbf{T}}
\newcommand{\dZ}{\mathbf{Z}}
\newcommand{\fI}{\mathfrak{I}}
\newcommand{\fm}{\mathfrak{m}}
\newcommand{\ft}{\mathfrak{t}}
\newcommand{\sF}{\mathscr{F}}
\newcommand{\arithfrob}{\mathrm{fr}}
\newcommand{\epic}{\twoheadrightarrow}
\newcommand{\etale}{\textnormal{\'et}}
\newcommand{\iso}{\xrightarrow\sim}
\newcommand{\mmu}{\boldsymbol\mu}
\newcommand{\mult}{\mathrm{m}}
\newcommand{\pow}[1]{\llbracket #1 \rrbracket}
%sha
\DeclareFontFamily{U}{wncy}{}
\DeclareFontShape{U}{wncy}{m}{n}{<->wncyr10}{}
\DeclareSymbolFont{mcy}{U}{wncy}{m}{n}
\DeclareMathSymbol{\sha}{\mathord}{mcy}{"58} 
\newtheorem{theorem}[subsection]{Theorem}

\title{Basic results in the deformation theory of Galois representations}
\author{Daniel Miller}

\begin{document}
\maketitle
\setcounter{tocdepth}{1}
\tableofcontents





This is a review of useful results in the study of deformations of (mostly 
two-dimensional) representations of $\pi_1(\dQ)=\galois(\overline\dQ/\dQ)$. 
References to the literature will be given whenever possible. 





\section{Group cohomology}


\subsection{Inflation-restriction}

This is from \cite[1.6.7]{nsw08}. Let $H\subset G$ be a closed normal subgroup 
of a profinite group. If $A$ is a $G$-module, then there is a canonical 
exact sequence 
\[\xymatrix{
  0 \ar[r] 
    & \h^1(G/H,A^H) \ar[r]^-{\mathrm{inf}} 
    & \h^1(G,A) \ar[r]^-{\mathrm{res}} 
    & \h^1(H,A)^{G/H} .
}\]


\subsection{Duality theorems for Galois cohomology}

Let $l$ be a prime, $X$ a connected noetherian scheme on which $l$ is 
invertible. Let $\dZ_l=\varprojlim\mmu_{l^n}$, considered as a smooth $l$-adic 
sheaf on $X$. For any $l$-adic sheaf $F$ on $X$, put 
$F(n)=F\otimes_{\dZ_l} \dZ_l(1)^{\otimes n}$. 

We call a \emph{$p$-adic field} a nonarchimedean local field of characteristic 
zero with residue characteristic $p$. 

\begin{theorem}[Tate]
Let $k$ be a $p$-adic local field. Let $M$ be a finite $\pi_1(k)$-module. Then 
the cup-product induces an isomorphism 
\[
  \h^\bullet(k,M^\vee(1)) = \h^{2-\bullet}(k,M)^\vee .
\]
\end{theorem}

Let $\pi=\pi_1(k)$, and let $M$ be a $\pi$-module. Suppose we want to compute 
$h^\bullet(M)$. It should be possible to compute 
$h^0(M)$ and $h^2(M)=h^0(M^\vee(1))$. We then use the Euler-Poincar\'e characteristic 
formula of Tate \cite[7.3.1]{nsw08} to do this. 


\subsection{Tate-Shafarevich groups and sets of places}

Let $F$ be a number field, $S$ a finite set of places of $F$. If $M$ is a 
$G_{F,S}$-module, put 
\[
  \sha_S^1(M) = \ker\left(\h^1(G_{F,S},M) \to \bigoplus_{v\in S} \h^1(G_v,M)\right) .
\]
If $S\subset T$, then one can naturally identify $\sha_S^1(M)$ with a subgroup of 
$\sha_T^1(M)$. Indeed, there is a natural injection (inflation) 
$\h^1(G_{F,S},M) \to \h^1(G_{F,T},M)$ coming from the projection 
$G_{F,T}\twoheadrightarrow G_{F,S}$. The five-term inflation-restriction exact 
sequence \cite[1.6.7]{nsw08} tells us that the image of the inflation map is 
the kernel of the restriction map $\h^1(G_{F,T},M) \to \h^1(H,M)$, where 
$H=\ker(G_{F,T}\to G_{F,S})$. The point is that 
$H=\langle I_v:v\in T\smallsetminus S\rangle$. So if $c\in \sha_T(M)$, then 
$c|_v=0$ for all $v\in T\smallsetminus S$, so certainly $c$ is induced from 
an element of $\h^1(G_{F,S},M)$. What remains is the easy verification of 
$c\in \sha_S(M)$. To be precise, 
$\sha_T^1(M)\subset \h^1(G_{F,T},M)$ is a subset of the image of 
$\sha_S^1(M)\subset \h^1(G_{F,S},M)$ under the (injective) inflation map. 





\section{Galois representations associated to modular forms}

Let $N\geqslant 1$ be an integer and $\varepsilon:(\dZ/N)^\times \to S^1$ a 
character. We write $S_0(N,\varepsilon)$ for the space of cusp forms for 
$\Gamma_1(N)$ with nebentypus $\varepsilon$. We call a form 
$f=\sum_{n\geqslant 0} a_n q^n$ in $S_0(N,\varepsilon)$ \emph{normalized} if 
$a_0=1$. 

\begin{theorem}
Let $N\geqslant 3$ and $k\geqslant 1$ be integers, $l$ an odd prime. Let 
$f_0\in S_0(N,\varepsilon)$ be a normalized eigenfunction for the Hecke 
operators $\{T_p:p\nmid N\}$. Let $K=K_f=\dQ(a_n:n\geqslant 1)$. 
Then there is a continuous irreducible representation 
$\rho_{f,l}:\pi_1\left(\dZ[\frac{1}{l N}]\right)\to \GL_2(K_{f,l})$ such that 
for each prime $p\nmid l N$, 
\begin{align*}
  \trace \rho_{f,l}(\arithfrob_p) &= a_p \\
  \det \rho_{f,l}(\arithfrob_p) &= \varepsilon(p) p^{k-1} .
\end{align*}
This representation is unique up to isomorphism. 
\end{theorem}
\begin{proof}
\textbf{Do this!}
\end{proof}





\section{Specific representations}

Nice fact if $\phi,\psi$ are characters: 
\[
  \adjoint(\phi\oplus \psi) = \phi^{-1}\psi\oplus \phi\psi^{-1}\oplus 2 .
\]
In particular, 
\[
  h^0(\adjoint(\phi\oplus 1)) = 2+2 h^0(\phi)
\]


\subsection{Peu ramifi\'ee and tr\`es ramif\'ee extensions}

The original source is \cite[2.4.6]{serre-1987}. 
Let $\bar\rho:G_{\dQ_p}\to \GL_2(\dF_q)$ be 
an ordinary representation, i.e. $\bar\rho$ is the extension of an unramified 
character by an unramified twist of the cyclotomic character. Let 
$\dQ_p^\mathrm{ur}(\bar\rho)$ be the extension of $\dQ_p^\mathrm{ur}$ with 
Galois group cut out by $\bar\rho(I)$, where $I\subset G_{\dQ_p}$ is the 
inertia group. It has a subextension $\dQ_p^\mathrm{ur}(\bar\rho|_P)$, where 
$P\subset I$ is wild inertia. Kummer theory tells us that 
\[
  \dQ_p^\mathrm{ur}(\bar\rho) = \dQ_p^\mathrm{ur}\left(\bar\rho|_P)(\sqrt[p]{x_1},\dots,\sqrt[p]{x_r}\right).
\]
We say that $\bar\rho$ is \emph{peu ramifi\'ee} if $v_p(x_i)\equiv 0\pmod p$ 
for each $i$, and $\bar\rho$ is \emph{tr\`es ramifi\'ee} otherwise. 

In \cite[8.2]{edixhoven-1992}, we have an alternative definition. Consider the 
extension $\bar\rho$ as a finite \'etale group scheme $V$ of $\dF_q$-vector 
spaces over $\dQ_p$. Then $\bar\rho$ is peu ramifi\'ee if $V$ can be extended 
to a finite flat group scheme over $\dZ_p$, and tr\`es ramifi\'ee otherwise. 


\subsection{Fundamental characters}

The reference is \cite[4.4]{tate-1995}.
Let $(\cO,\fm,k)$ be a complete mixed-characteristic discrete valuation ring 
with perfect residue field. Then the projection $\cO\twoheadrightarrow k$ 
admits a multiplicative section $\omega:k\to \cO$. If $k_0$ is a field, then 
the induced map $k_0\to \cO$ coming from any embedding $k_0\hookrightarrow k$ 
is called a \emph{fundamental character}. The main example is when $\cO$ is the 
ring of integers in a finite extension of $\dQ_p$ and 
$k=\dF_{p^f}$, in which case the fundamental characters 
$k^\times\to \cO^\times$ form a $\dZ/f$-torsor under 
$r\cdot \chi = \chi^{p^r}$. 

A better reference is \cite[1.7]{serre-1972}. 





\section{Modular representations}


\subsection{Hecke operators}

A good (concise) summary of the diamond operators, Atkin-Lehner involution, and 
Hecke operators is \cite[ch.2 \S 5]{mw84}. 


\subsection{New parts of Jacobians}

The following is from \cite[\S 2]{mazur-1978}. 
For $n\geqslant 1$, let $J_0(n)$ be the jacobian of the modular curve $X_0(n)$. 
If $n=n' d$, there is a ``degeneracy map'' $B_d:X_0(n)\to X_0(n')$ that sends a 
pair $(E,C)$ consisting of an elliptic curve and $C\subset E[n]$ of order $n$ 
to the pair $(E/C[d],(C/C[d])[n'])$. There are induced maps 
$B_d^\ast:J_0(n')\to J_0(n)$. Let $J_0(n)_\mathrm{old}\subset J_0(n)$ be the 
abelian subvariety generated by the images of the $B_d$ for $n'<n$, and define 
$J_0(n)^\mathrm{new}$ by the short exact sequence 
\[
  0 \to J_0(n)_\mathrm{old} \to J_0(n) \to J_0(n)^\mathrm{new} \to 0 .
\]
By general theory, there is an isogeny 
$J_0(n)\sim J_0(n)_\mathrm{old}\times J_0(n)^\mathrm{new}$, thus an isomorphism 
of Galois representations 
\[
  V_\ell J_0(n) \simeq V_\ell J_0(n)_\mathrm{old}\oplus V_\ell J_0(n)^\mathrm{new} .
\]
There is an induced action of the Hecke algebra on $J_0(n)^\mathrm{new}$. 


\subsection{Eisenstein ideal}

This definition is from \cite[II.9]{mazur-1977}. Let $\dT=\dT_n$ be the Hecke 
algebra for $\Gamma_0(n)$. So $\dT$ is generated as a $\dZ$-algebra by the 
Hecke operators $T_l$, $l\nmid n$. The \emph{Eisenstein ideal} 
$\fI\subset \dT$ is generated by the $T_l-(l+1)$ for $l\nmid n$, and 
$1+w$. So if $f\in S_k$ is an eigenform annihilated by $\fI$, one has 
$a_p(f) = p+1$. This means $\rho_{f,l}$ should look like 
$\kappa_l\oplus 1$, where $\kappa$ is the cyclotomic character. 





\section{Deformation problems}

Let $\cO$ be a complete dvr with residue field $k$. Our deformation problems 
will be covariant functors on the category $\mathsf{C}_\cO$ of ``test 
objects.'' These are local artinian $\cO$-algebras $A$ such that 
$\cO\to A$ induces an isomorphism $k\iso A/\fm_A$. 


\subsection{Minimal deformations}

Here we follow \cite[\S2.1]{khare-2003}. Let $k$ be a finite field of 
characteristic $p$ and $\bar\rho:G_{\dQ,S}\to \GL_2(k)$ a continuous 
$p$-ordinary representation. One says a lift $\rho:G_{\dQ,S}\to \GL_2(A)$ is 
\emph{minimally ramified} if for $v\in S\smallsetminus p$, 
\[
  \rho|_{I_v}\sim \begin{pmatrix} 1 & \ast \\ & 1 \end{pmatrix} .
\]
\textsf{(This doesn't seem to be the same as 
\cite[p.180]{khare-ramakrishna-2003}. Find out what's wrong.)}


\subsection{New deformation rings}

We follow \cite[df.1]{khare-ramakrishna-2003}. 
Let $\bar\rho:G_\dQ\to \GL_2(\dF_q)$ be a continuous representation unramified 
outside $S$. Suppose $T\supset S$ is a finite set of primes such that 
$\bar\rho$ is nice for $T\smallsetminus S$. Then 
$R_{\bar\rho}^{T\textnormal{-new}}$ represents minimally ramified deformations 
$\rho:G_{\dQ,S} \to \GL_2(A)$ such that for $v\in T\smallsetminus S$, 
$\rho_v$ is a twist of $\begin{pmatrix} \varepsilon & \ast \\ & 1 \end{pmatrix}$. 





\section{Commutative algebra}


\subsection{Weierstrass preparation theorem}

This is from \cite[VII \S 3.8, pr.6]{bourbaki-comm-alg}. Let $\cO$ be a 
complete discrete valuation ring with uniformizer $\pi$. Then any 
$f\in \cO\llbracket X\rrbracket$ can be written as 
\[
  f = u \pi^m \left(X^n+a_{n-1} X^{n-1} + \cdots + a_0\right),
\]
where $u\in \cO\llbracket X\rrbracket^\times$ and the 
$a_i\in \langle \pi\rangle$. In particular, the only way the quotient 
$\cO\llbracket X\rrbracket/f$ can be flat over $\cO$ is for 
$m=0$, in which case the quotient has finite $\cO$-rank. 


\subsection{Specific presentations via small extensions}

Fix a finite field $k$ of characteristic $p$. Recall that a \emph{coefficient 
ring} over $k$ is a complete local noetherian $\witt(k)$-algebra with residue 
field $k$. If $R$ is such a ring, write $\ft_R=\hom(\fm_R/\fm_R^2,k)$; this is 
a $k$-vector space. Recall that a \emph{small extension} of coefficient rings 
over $k$ is a surjection $R_1\epic R_0$ such that the kernel $I$ is principle 
and annihilated by $\fm_1$.

We are interested in measuring the complexity of presentations of coefficient 
rings. Write $\witt(k)\pow{\bm x}=\witt(k)\pow{x_1,\dots,x_d}$. For a 
polynomial $f\in \witt(k)\pow{\bm x}$, put 
\[
  v(f) = \min\{e:p^e\mid f\} + \sum_{i=1}^r \min\{n_i:x_i^{n_i}\mid f\} .
\]
For a set $\bm f=\{f_1,\dots,f_r\}\subset \witt(k)\pow{\bm x}$, the 
\emph{complexity} of $\bm f$, denoted $v(\bm f)$, is by definition 
$\min\{v(f_i)\}_{1\leqslant i\leqslant r}$. Put $|\bm n|=n_1+\cdots+n_r$. 
Note that if $v(\bm f)\geqslant e+|\bm n|$, then we have a surjection 
\[
  R(e,\bm n) = \witt(k)\pow{\bm x}/\langle p^e,x_1^{n_1},\dots,x_d^{n_d}\rangle \epic R(\bm f) = \witt(k)\pow{\bm x}/\langle f_1,\dots,f_r\rangle .
\]

We introduce an operation $\bm f\mapsto \bm f^+$ on sets of relations. Put 
\[
  \{f_1,\dots,f_r\}^+ = \{p f_1,x_1 f_1,\dots,x_d f_1,\dots,p f_r,x_1 f_r,\dots,x_d f_r\} .
\]
Note that $v(\bm f^+)>v(\bm f)$, and that the natural map 
$R(\bm f^+)\epic R(\bm f)$ factors as 
\begin{align*}
  R(\bm f^+) 
    &\epic R(p f_1,x_1 f_1,\dots,x_d f_1,\dots,p f_{r-1},x_1 f_{r-1},\dots,x_d f_{r-1},f_r) \\
    &\epic R(p f_1,x_1 f_1,\dots,x_d f_1,\dots,p f_{r-2},x_1 f_{r-2},\dots,x_d f_{r-2},f_{r-1},f_r) \\
    &\epic \cdots \\
    &\epic R(\bm f) ,
\end{align*}
in which each surjection is small. 

Write $\bm f^{+0}=\bm f$, $\bm f^{+(n+1)}=(\bm f^{+n})^+$. Fix some 
$\bm f$. Then for all $(e,\bm n)$ with $e+|\bm n|\geqslant v(\bm f)$, there 
exists some $m$ such that $v(\bm f^{+m})\geqslant e+|\bm n|$. This gives quotients 
\[
  R(e,\bm n)\twoheadleftarrow R(\bm f^{+m})\epic R(\bm f) .
\]
The key facts here are:
\begin{enumerate}
  \item The surjection $R(\bm f^{+m})\epic R(\bm f)$ is a composite of small 
    extensions. 
  \item Rings of the form $R(e,\bm n)$ surject onto any finite coefficient ring. 
\end{enumerate}
The latter fact holds because $\witt(k)\pow{\bm x}=\varprojlim R(e,\bm n)$.





\bibliographystyle{alpha}
\bibliography{tidbit-sources}

\end{document}
