\documentclass[handout]{beamer}
% Remove [handout] for pause to work.
\usetheme{metropolis}

\DeclareMathOperator{\cdf}{cdf}
\DeclareMathOperator{\D}{D}
\DeclareMathOperator{\GL}{GL}
\DeclareMathOperator{\N}{N}
\DeclareMathOperator{\R}{R}
\DeclareMathOperator{\ram}{ram}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\ST}{ST}
\DeclareMathOperator{\SU}{SU}
\DeclareMathOperator{\sym}{sym}
\DeclareMathOperator{\tr}{tr}
\DeclareMathOperator{\X}{X}
\newcommand{\bC}{\mathbf{C}}
\newcommand{\bF}{\mathbf{F}}
\newcommand{\bQ}{\mathbf{Q}}
\newcommand{\bR}{\mathbf{R}}
\newcommand{\bZ}{\mathbf{Z}}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\frob}{\mathrm{fr}}

\newtheorem{conjecture}{Conjecture}

\title{Counterexamples related to the Sato--Tate conjecture}
\author{Daniel Miller}
\institute{Cornell University}
\date{25 April 2017}





% Outline:
% Motivation: 
%  - What is the problem I'm trying to solve?
%    2 problems: does RH => AT? is there a Galois proof of ST?
%  - Why is this problem interesting or important?
%    * AT=>RH: so far proofs of ST have proved things about L-functions. 
%      How far can knowledge of L-functions take you?
%    * Galois ST: does ST conjecture follow from Galois theory alone? 
%  - Why is my solution interesting?
%    * Two "No!" answers. 
%    * Proving AT will take more than just L-functions.
%    * ST fundamentally involves automorphy!
% Work
%  - What theorems have I proved? 
%    * "Master Galois theorem"
%    * "dioph approx counterexample"
%  - What systems have I built? 
%    * see above. 
% Related work
%  - Who else has attacked this problem or similar problems? 
%    * Pande, KLR, Thorner, Mazur
%  - In what ways is my solution better? 
%    * Better than Pande: stronger results
%    * convenient "packaging" of KLR
%    * shows that Thorner doesn't generalize to CM
%  - In what ways is it limited? 
%    * only addresses CM varieties - may not work for SU(2)
% So what?
%  - need new directions to prove AT, even for CM case. 





\begin{document}

\begin{frame}
\titlepage
\end{frame}

\begin{frame}
\frametitle{Outline}
\tableofcontents
\end{frame}





\section{Motivation}

\begin{frame}{Sato--Tate Conjecture}
$E_{/\bQ}$ non-CM elliptic curve, $a_p = \tr\rho_l(\frob_p)\in \bZ$, 
$|a_p|\leqslant 2\sqrt p$. 
\pause

Satake parameter: $\theta_p = \cos^{-1}\left( \frac{a_p}{2\sqrt p}\right)$. 
\pause

Sato--Tate measure: $\ST = \frac{2}{\pi} \sin^2 \theta\, \dd \theta$ (Haar 
measure on $\SU(2)^\natural$). 
\pause

\begin{theorem}[Taylor et.~al.]
$\{\theta_p\}$ is equidistributed with respect to $\ST$. 
\end{theorem}
\pause

Quantify rate of convergence of 
$\frac{1}{\pi(N)} \sum_{p\leqslant N} \delta_{\theta_p}$ to $\ST$. 
\pause

Use discrepancy (Kolmogorov--Smirnov statistic). 
\end{frame}


\begin{frame}{Akiyama--Tanigawa Conjecture}
\[
	D_N = \sup_{x\in [0,\pi]}\left| \frac{1}{\pi(N)} \sum_{p\leqslant N} 1_{[0,x)}(\theta_p) - \int_0^x \, \dd\ST\right| .
\]
\pause

\begin{conjecture}[Akiyama--Tanigawa]
$D_N \ll N^{-\frac 1 2 + \epsilon}$. 
\end{conjecture}
\pause

There is a variant of this conjecture for CM elliptic curve.
\pause

\begin{theorem}[Akiyama--Tanigawa]
Akiyama--Tanigawa conjecture $\Rightarrow$ Riemann Hypothesis for $E$.
\end{theorem}
\pause

\begin{theorem}[Mazur]
Akiyama--Tanigawa conjecture $\Rightarrow$ Riemann Hypothesis for $\sym^k E$
\end{theorem}
\end{frame}


\begin{frame}{Context}
Pande: is the Sato--Tate conjecture a Galois-theoretic result?
\pause

\begin{theorem}[Pande]
Let $\epsilon>0$. Then there exists an infinitely ramified representation 
$\rho\colon G_\bQ \to \GL_2(\bZ_l)$ such that $\theta_p\in B_\epsilon(\pi/2)$ 
for a density one set of primes. 
\end{theorem}
\pause

\begin{theorem}[Khare--Rajan]
Any $\rho\colon G_\bQ \to \GL_2(\bZ_l)$ is ramified at a density zero set of 
primes. 
\end{theorem}
\pause

\textbf{Question (Serre):} can you \emph{a priori} control the ratio 
$\frac{\pi_{\ram(\rho)}(x)}{\pi(x)}$?
\pause

\textbf{Answer (Khare--Larsen--Ramakrishna):} no!
\end{frame}


\begin{frame}{Questions}
\textbf{Q1.} Can Pande's results be strengthened to yield equidistribution? 
\pause

\textbf{Q2.} If so, can the measure be specified?
\pause

\textbf{Q3.} Can the rate of convergence of empirical measures to the true measure 
be specified?

\textbf{Q4.} Can the ratio $\frac{\pi_{\ram(\rho)}(x)}{\pi(x)}$ be controlled?
\pause

\textbf{Q5.} Can anything be said about the $L$-functions associated with $\rho$?
\pause

\textbf{Answer:} Yes! to Q1--Q5. 
\end{frame}





\section{Discrepancy and Dirichlet series}


\begin{frame}{Discrepancy}
\begin{definition}
Let $\{\theta_p\}$ be a sequence in $[0,\pi]$, $\mu$ a measure on $[0,\pi]$. 
The \emph{discrepancy} is 
\[
	\D_N(\{\theta_p\},\mu) =  \sup_{x\in [0,\pi]}\left| \frac{1}{\pi(N)} \sum_{p\leqslant N} 1_{[0,x)}(\theta_p) - \int_0^x \, \dd\mu\right| .
\]
\end{definition}
\pause

$\{\theta_p\}$ are $\mu$-equidistributed if and only if $D_N \to 0$. 
\pause

\textbf{Fact:}
$\frac{\log N}{N} \ll D_N$. The \emph{van der Corput sequence} achieves this. 
\end{frame}


\begin{frame}{Dirichlet series}
\begin{definition}
For $k\geqslant 1$, 
\[
	L(\sym^k \rho,s) = \prod_p \det\left( 1 - \sym^k \begin{pmatrix} e^{i \theta_p} & 0\\ 0 & e^{-i \theta_p}\end{pmatrix} p^{-s}\right)^{-1}
\]
\end{definition}
\pause

\begin{definition}
For $f\colon [0,\pi] \to \bC$ of bounded variation with $\mu(f) = 0$,  
\[
	L_f(s) = \prod_p \left( 1 - f(\theta_p) p^{-s}\right)^{-1}
\]
\end{definition}
\pause

\textbf{Example:} $L_{\sgn}(s) = \prod_p \left(1 - \sgn(a_p) p^{-s}\right)^{-1}$. 
\end{frame}

\begin{frame}{Dirichlet series---results}
\begin{theorem}
If $\left| \sum_{p\leqslant N} f(\theta_p)\right| \ll N^{\alpha+\epsilon}$, 
then $L_f(s)$ admits a nonvanishing analytic continuation to $\Re > \alpha$. 
\end{theorem}
\pause

\begin{corollary}
If $D_N \ll N^{-\alpha+\epsilon}$, then $L_f(s)$ admits a nonvanishing 
analytic continuation to $\Re > \alpha$. 
\end{corollary}
\pause

\begin{definition}
$U_k(\theta) = \frac{\sin((k+1)\theta)}{\sin\theta} = \tr\sym^k \left(\begin{smallmatrix} e^{i \theta_p} & 0\\ 0 & e^{-i \theta_p}\end{smallmatrix}\right)$.
\end{definition}
\pause

\begin{theorem}
If $\left| \sum_{p\leqslant N} U_k(\theta_p)\right| \ll N^{\alpha+\epsilon}$, 
then $L(\sym^k\rho, s)$ admits a nonvanishing analytic continuation to 
$\Re > \alpha$. 
\end{theorem}
\end{frame}





\section{Main theorem}


\begin{frame}{Ingredients}
1. Fix a rational prime $l\geqslant 7$. 
\pause

2. Fix an odd, absolutely, weight 2 representation 
$\bar\rho\colon G_\bQ \to \GL_2(\bF_l)$.
\pause
\textcolor{red}{$\rho$ will be a lift of $\bar\rho$.}
\pause

3. Fix a function $h\colon \bR^+ \to \bR_{\geqslant 1}$ which increases slowly 
to infinity. 
\pause
\textcolor{red}{We will have $\pi_{\ram(\rho)}(x) \ll h(x)$.}
\pause

4. Fix an absolutely continuous measure $\mu$ on $[0,\pi]$, with bounded 
probability density function.
\pause
\textcolor{red}{The angles $\{\theta_p\}$ will be $\mu$-equidistributed.}
\pause

5. Fix $\alpha\in \left(0,\frac 1 2\right)$. 
\pause
\textcolor{red}{The discrepancy $D_N$ will decay like $\pi(N)^{-\alpha}$.}
\end{frame}


\begin{frame}{Theorem}
Let $l$, $\bar\rho$, $h$, $\mu$, and $\alpha$ be as above. Then there exists 
$\rho\colon G_\bQ \to \GL_2(\bZ_l)$ such that 
\begin{enumerate}
\item
$\rho \equiv \bar\rho\pmod{l}$. 
\pause

\item
$\pi_{\ram(\rho)}(x) \ll h(x)$. (Yes to Q4.)
\pause

\item
For each unramified $p$, $a_p = \tr \rho(\frob_p)\in \bZ$ and satisfies the 
Hasse bound.
\pause

\item
$\D_N(\{\theta_p\},\mu) = \Theta(\pi(x)^{-\alpha})$. (Yes to Q1--Q3.)
\pause

\item
If $(\theta\mapsto \pi-\theta)_\ast \mu = \mu$, then for each odd $k$, 
$L(\sym^k \rho,s)$ satisfies the Riemann hypothesis. (Yes to Q5.)
\end{enumerate}
\end{frame}






\section{Idea of the proof}


\begin{frame}{Prescribing discrepancy decay}
\begin{theorem}
If $\alpha\in \left(0,\frac 1 2\right)$, there exists a sequence 
$(x_2,x_3,x_5,\dots)$ in $[-1,1]$ such that 
$| D_N - \pi(N)^{-\alpha}| \ll \pi(N)^{-1}$. 
\end{theorem}
\pause

\textbf{Fact:} 
Discrepancy is invariant under pushforward by $\cos$ and $\cos^{-1}$. 
\pause

\textbf{Idea}:
Construct $\rho$ so that $\frac{a_p}{2\sqrt p}\approx x_p$. 
\pause

\textbf{Fact:} 
If $(x_p^{(1)})$ is a sequence with $|x_p - x_p^{(1)}| \ll p^{-1/2+\epsilon}$, 
then $\D_N^{(1)} = \Theta(\pi(N)^{-\alpha})$. 
\end{frame}


\begin{frame}{Lifting Galois representations}
What is a deformation?

Weight?

What does it take to lift from $\bZ/l^n$ to $\bZ/l^{n+1}$?
\end{frame}


\begin{frame}{Lifting Galois representations---first stage}
Lift from $\bZ/l$ to $\bZ/l^2$. 
\end{frame}


\begin{frame}{Lifting Galois representations---inductive step}
Lift from $\bZ/l^n$ to $\bZ/l^{n+1}$. 
\end{frame}


\begin{frame}{Consequences}
If $f\in C([0,\pi])$, $f\circ \cos^{-1}\colon [-1,1]\to \bC$ is Lipschitz, and 
$f(\pi-\theta) = - f(\theta)$, then $L_f(\rho,s)$ has a nonvanishing analytic 
continuation to $\Re > \frac 1 2$ (Riemann hypothesis). 
\end{frame}





\begin{frame}
\begin{center}
\Huge Questions?
\end{center}
\end{frame}





\end{document}
